\documentclass[a4paper]{article}

\usepackage[top=2cm, bottom=2cm, left=3cm, right=3cm]{geometry}

\usepackage{float}

\usepackage{multirow}

\usepackage[hyphens]{url}
\usepackage{hyperref}

\usepackage{appendix}
\usepackage[numbers]{natbib}

\usepackage{graphicx}

\usepackage{amsmath}

\usepackage[parfill]{parskip}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.75\baselineskip}

\newcommand{\mytilde}{\raise.17ex\hbox{$\scriptstyle\mathtt{\sim}$} }

\title{CS407 - Project Specification:\\
Towards Debugging Wireless Sensor Network Applications}
\date{25th October 2012}
\author{
	Matthew Bradbury (0921660) \and
	Tim Law (0918647) \and
	Ivan Leong (0830934) \and
	Daniel Robertson (0910210) \and
	Amit Shah (0904778) \and
	Joe Yarnall (0905247)
}

\begin{document}

\maketitle

%No numbering on first page
\pagestyle{empty}
\thispagestyle{empty}

\newpage

\pagestyle{plain}
\setcounter{page}{1}

\tableofcontents
\clearpage


\section{Introduction}
The recent emergence of Wireless Sensor Networks (WSNs) due to their decrease in
cost, size and increase in battery life has seen their rising application in
many real world problems \cite{wsnapps}. These applications fall into 3
categories: Military Applications - general information collection such as enemy
tracking, battlefield surveillance or target classification are all areas being
researched actively, Environmental Monitoring - both indoor monitoring for
emergency services and outdoor monitoring with applications to monitoring
ecology systems and agricultural processes and Logistics - ranging applications
from asset tracking using RFIDs to monitoring a company's industrial process and
heavy machinery. As can be clearly seen the applications for WSNs is vast, as
such it has become a very active research area.

However, when working with WSNs there are several considerations that a
developer must make. Typical wireless protocols are not practically
applicable in these networks, due to a set of different constraints, such as
energy consumptions and a limited wireless range. This presents new challenges
when designing algorithms and software for these devices. In an attempt to solve
these problems software will need to be developed to operate the WSN hardware,
and like all software this software will contain bugs \cite{5010224}. As sensor
networks are a distributed system, this increases the classes of bugs that
software can suffer from \cite{5010224}, such as dead-lock, live-lock and other
concurrency issues. To aid in developing reliable software for WSNs, there is
thus need for tools that can assist in detecting and reporting these faults, so
a developer can correct them and create reliable software.

\section{Motivation}
The need for debugging has been around for as long as there has been hardware
that performs a task. With a famous folklore example being the case where the
term debugging was coined, where a moth was picked out of a crucial relay
\cite{shapiro1987etymology}. However, nowadays debugging searches for much more
complicated problems than moths in machines. This is particularly apparent with
WSNs which is a relatively new field, and as a result there is a limited number
of tools available for debugging WSNs, despite large increases in the research
into communication protocols \cite{TankBible}. 

When you have a sequential program it is easy to insert breakpoints or debug
messages that allow you to inspect the state of the program at any given point
in its execution. However, with a distributed system, they can have the same
issues that sequential programs have, but they also have the possibility of
bugs related with the interaction of each of the different nodes in the system.
These interactions can be non-deterministic \cite{liu2007wids} in nature and
this can make inspection of the system's global state at any given point during
execution more challenging \cite{chandy1985distributed}. These links can lead to
many types of bugs that a sequential program would never encounter, such as
deadlock \cite{singhal1989deadlock}, inconsistent states
\cite{chandy1985distributed} and the need to detect and handle failures in other
nodes of the system \cite{5623434}. While there exist algorithms to solve some
of these problems (for instance TCP provides stronger guarantees of message
deliver than alternative protocols such as UDP
\cite[p.~532]{Tanenbaum:2002:CN:572404}), there are some errors that a WSN
developer will need to ensure they do not make when writing their software.

Wireless Sensor Networks are a class of distributed system, with a special
property that they are energy constrained \cite{6023235}. As they are a
distributed system, the same distributed bugs can occur in software developed
for them, therefore there is a need for tools to assist with this kind of
debugging. However, as they are energy constrained developers will attempt to
make certain optimisations that may not have been required with other types of
distributed systems. This means any tool that is specific to them will need to
be able to detect these kinds of bugs and relay them to the developer in a
manner which is both accurate and rich enough to be useful for debugging, yet 
concise enough to not overtax the nodes or batteries.

It is apparent to us that their is a need to create tools for WSN developers,
so that they can focus their time on developing applications rather than on
debugging software to fix them. These tools need to be able to provide
information to developers on errors that occur during the execution of their code 
when deployed to a WSN, as well as provide a visualisation of activity within a WSN.

\section{Project Description}

\subsection{Aims and Objectives}
\begin{itemize}
	\item An application that can check to see if for a certain subsection
of the network a given predicate holds.
	\item A visualisation tool for logged network traffic.
	\item Implementations of various distributed algorithms to test the
visualizer and the predicate checker.
\end{itemize}

\subsection{Visualisation Tool}
\begin{itemize}
	\item Visualisation of the sensor network with node positioning and
connections.
	\item Visualisation of active clusters within the sensor network.
	\item Visualisation of messages flowing through the network, along with
node logs of the information processed.
\end{itemize}

\subsection{Predicate Checker}
\begin{itemize}
	\item Functionality to query nodes to see if a given predicate holds.
	\item Visualisation of nodes that are breaking a given predicate during
runtime.
\end{itemize}

\subsubsection{Predicate Specification}

There are also several different ways that we can specify a predicate. The first
is simply writing a function that takes some input state and produce a boolean
output. We aim to produce a C library that will take this predicate, give it
some state and check the result. If the result is false then the library will
generate a debug message, containing logs of recent events and details of why
the predicate failed, then send that message back to the base station.

\begin{figure}[H]
\begin{center}
    \begin{displaymath}
	CheckTemp(T) = T > 0 \land T < 40
	\end{displaymath}
	\caption{For example we would like to check that the temperature sensor
is reporting values within the range its specification says it should produce}
\end{center}
\end{figure}

Another way that predicates can be defined is at run-time instead of at
compile-time. The idea here is that we would send a predicate written in a
scripting language to each of the nodes and then have that predicate interpreted
and executed at a given interval or at a certain place. The code running on the
nodes will need to register what memory locations correspond to which variable
so that the interpreter can give named variables values. We expect that this may
be difficult to do, due to the constrained resources of the node \cite{CM5000}.
However, there is a precedence that this is possible, the fact that there exists
TinyDB \cite{TinyDB} which would need to interpret SQL statements and then execute
them.

As for the type of logic that we intend to use, for compile-time predicate
checking the full expressiveness of C is available. For run-time checking there
may need to be restrictions on the type of logic we can use. Initially we are
aiming to use predicate logic with at least the ability to execute $\land$,
$\lor$, $\lnot$, $=$, $\neq$, $<$, $>$, $\leq$ and $\geq$ on basic C types such
as floating point variables and integers. If possible we would like to implement
quantifiers ($\forall$ and $\exists$) as we expect these to be useful for
network predicates. As for capabilities such as sets and function definitions
and data structure definitions, these may be too expensive and not supported.
Although we intend to investigate the use of a lightweight implementation of LUA
\cite{elua} or the Contiki Shell \cite{?} to evaluate these predicates, so the
features that we can support will relate to what these tools are capable of.

\begin{figure}[H]
\begin{center}
    \begin{displaymath}
	NeighbourhoodHealty() = \forall Node \in ExpectedNeighbourhood : NotCrashed(Node)
	\end{displaymath}
	\caption{Check that all nodes in the expected 1-hop neighbourhood have
not crashed}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
    \begin{displaymath}
	CheckAggregrationTreeHeight(Msg) = Msg.HopCount = ExpectedDepth
	\end{displaymath}
	\caption{Check that on receiving a message the height of the tree is as
expected. This aims to catch when a tree has been set up and the structure has
now changed, making the current structure suboptimal in energy usage.}
\end{center}
\end{figure}

\subsubsection{Predicate Checking}

There are several types of predicates that we may wish to check. The first kind
are very simple predicates that one might check if a standard sequential
non-distributed program, such as when performing an \verb|ASSERT| statement or
checking a condition then throwing an exception. Here we would use the library
we talked about developing above, and place these checks into the code at
desired places.

The next level of checking would be checking of predicates that specify some
invariant of the program. These could be hard-coded into the software or
provided a run-time. These could be run at given points throughout the execution
(at beginning and end of functions) or randomly throughout the program's
execution.

The final level of checking involves network level predicates. These predicates
would check for things such as permanent node failures and also be able to
recognise temporary node failures. Other examples where they might be used are
to check that nodes are obeying the TDMA (Time Division Multiple Access) slot
they have been allocated. Or if an aggregation tree \cite{TankBible} that has
been set up for a WSN is still optimal (for instance checking that the height of
the tree is the same as the maximum hop count of messages received at the base
station).

This final network checking has some additional considerations to be made. For
instance if we take mobile nodes into account then detecting crashed nodes
becomes more difficult. With stationary nodes you can just check that your 1-hop
neighbourhood remains the same. With mobile nodes you expect the 1-hop
neighbourhood to change, so will have to use a different strategy.


\subsubsection{Algorithms to Implement}

We plan on implementing two sets of algorithms, the first are those that will
report the predicate failures. The following are the algorithms we intend to
implement.

\begin{itemize}
	\item H-SEND \cite{herbert2007adaptive}
	\item Our own implementation drawing on ``Send to Base'' style techniques
\end{itemize}

The second set of algorithms we intend to implement are those that we will use
to test the algorithms that check that given predicates hold.

\begin{itemize}
	\item Clustering algorithm (such as LEACH) \cite{herbert2007adaptive}.
	\item An algorithm to set up an aggregation tree (for some information
like temperature average). \cite{TankBible}
\end{itemize}

\subsection{Required Resources}
\begin{itemize}
	\item Wireless Sensor Nodes and related interface hardware
	\item Operating System and Simulator for the Wireless Sensor Nodes
	\item Laptop computers to act as mobile base stations, and for development
	\item Access to academic papers that detail the previous work
	\item A DVCS where we can store source code for the project
	\item A simulator and development environment that will allow us to test code locally, then deploy to the actual hardware. There is a choice between Contiki \cite{23839452} and TinyOS \cite{levis2003tossim}
	\item If possible, the ability to develop code on the DCS lab machines
\end{itemize}

\section{Project Management}

\subsection{Management}

As this is a group project there are additional challenges to face compared to a
solo project. Therefore, we plan to use the knowledge and experience gained in
the module IB382 Project Management \cite{IB382} as well as our own experience
to avoid common problems that could prevent us working effectively. We also plan
to refer to the Project Management textbook for additional resources
\cite{PMTextBook}.

As part of the management we will make sure we meet up at least once a week to
discuss our current progress, and allocate tasks for the next week. We also
intent to stay in contact and discuss our progress and any resources we find in
between these meetings.

We will also be assigning roles, in order to keep the group organised. The
following table shows how we assigned roles, although we intend for these to be
flexible:

\begin{table}[H]
\centering
	\begin{tabular}{| c | c |}
		\hline
		Name & Role\\
		\hline
		Matthew Bradbury & Group Leader\\
		Tim Law & Developer and Researcher\\
		Ivan Leong & Developer and Tester\\
		Daniel Robertson & Project Manager\\
		Amit Shah & Technical Leader\\
		Joe Yarnall & Developer and Tester\\
		\hline
	\end{tabular}
\end{table}



\subsection{Development Methodologies}
In our first meeting with our project supervisor, we were given the objective to
develop a wireless sensor network (WSN) with the new set of sensors that was
purchased by the department of Computer Science. He proposed several ideas that
we could use to implement the system (e.g. routing protocols and algorithms) and
a few topics on how we can investigate the problems associated with WSN such as
global predicate detection and network debugging. 

For our project, we plan to follow a spiral evolutionary process model since we
are expecting many changes in our system, for example trying to implement the
best algorithm for predicate checking and experimenting with different routing
protocols. The first circuit around the spiral will involve the development of
the core functionalities and each evolutionary step may be used to build
prototypes and more complex versions of the system. Each pass through the spiral
may require our work to be reviewed by our supervisor and any adjustments to the
project plan will be made in accordance to the feedback. With prototyping
aspects of the model, it may help us reduce the chances of any risks.



\subsection{Schedule}

\subsubsection{Term 1}

During the first term, we plan to do our initial research around the subject to
find out what has been done and what is implementable for us. We then plan to
implement some of the algorithms to detect predicate failures we researched in a
simulator. Once the algorithms are working in a simulator we will deploy them to
the motes. We expect difficulty deploying to the motes due to: i) us needing to
learn how to do it, ii) the environment being slightly different to the
simulators and iii) debugging our algorithm will be much harder on the hardware
than in a simulator. By the end of week 9 we should have a basic implementation
for alerting when application predicates do not hold for stationary nodes.

\begin{table}[H]
	\centering
	\begin{tabular}{| l | l | l | l | l | l | l |}
	Task Description & \multicolumn{6}{|l|}{Time Allocated (Weeks)}\\
	~ & Amit & Dan & Ivan & Joe & Matt & Tim \\
	\hline
	\hline
	\multicolumn{7}{|l|}{\textbf{Term 1} - Developing for application predicate checking} \\
	\hline


	Research around the Problem & 2 & 2 & 2 & 2 & 2 & 2\\
	Writing Specification & 1 & 1 & 1 & 1 & 1 & 1\\
	Algorithm Development & 3 & 3 & 3 & 3 & 3 & 3\\
	H-SEND Implementation & ~ & ~ & ~ & ~ & ~ & ~\\
	``Send to Base'' Implementation & ~ & ~ & ~ & ~ & ~ & ~\\
	Clustering Implementation & ~ & ~ & ~ & ~ & ~ & ~\\
	Aggregation Tree Implementation & ~ & ~ & ~ & ~ & ~ & ~\\
	Testing & 1 & 1 & 1 & 1 & 1 & 1\\
	Testing and adapting to Physical Nodes & 2 & 2 & 2 & 2 & 2 & 2\\
	Poster Creation and Presentation preparation & 1 & 1 & 1 & 1 & 1 & 1\\

	\hline
	
	\end{tabular}
\end{table}

\subsubsection{Term 2}

We do not expect to fully finish development for application predicates during
the first term, so we will continue development during the second term. We also
aim to research and implement different algorithms focusing on different aspects
of wireless sensor networks (such as node mobility and checking network
predicates). Again a lot of time is devoted to testing and deployment due to the
difficulties involved.

\begin{table}[H]
	\centering
	\begin{tabular}{| l | l | l | l | l | l | l |}
	Task Description & \multicolumn{6}{|l|}{Time Allocated (Weeks)}\\
	~ & Amit & Dan & Ivan & Joe & Matt & Tim \\
	\hline
	\hline
	\multicolumn{7}{|l|}{\textbf{Term 2} - Developing for network predicate checking (and mobile nodes)} \\
	\hline
	
	Additional Research & 1 & 1 & 1 & 1 & 1 & 1\\
	Algorithm Development & 3 & 3 & 3 & 3 & 3 & 3\\
	Adapt our ``Send to Base'' Implementation & ~ & ~ & ~ & ~ & ~ & ~\\
	Testing & 2 & 2 & 2 & 2 & 2 & 2\\
	Testing and adapting to Physical Nodes & 2 & 2 & 2 & 2 & 2 & 2\\
	Report Writing & 2 & 2 & 2 & 2 & 2 & 2\\
	
	\hline
	
	\end{tabular}
\end{table}


\section{Potential Risks}

Throughout our project we expect to encounter difficulties and challenges. The
following are the risks that we think may be encountered during the project and
ways that these risks could be mitigated.


\subsection{Risks}
\begin{itemize}
	\item Needing to talk to our project supervisor (Arshad), even though he is on sabbatical.
	\item Project drifting in the wrong direction due to uncertainty.
	\item Availability of hardware, because it is shared by the department.
	\item Lack of features in the simulator and/or wireless sensor network environment.
	\item Problems that we have not predicted here.
	\item Team members being unavailable to do work (eg. due to illness).
\end{itemize}

\subsection{Risk Mitigation}
\begin{itemize}
	\item Even though Arshad is on sabbatical he should be in the department
3 days a week. Also there are other departmental staff that can assist us if
Arshad is not available.
	\item If our project is not heading in the right direction, we will seek
guidance from our supervisor who should be able to guide us onto the right
track.
	\item Additionally to stop project drift we will have weekly meetings to
assess our progress and set clear goals for each member of the group.
	\item If our development environment lacks features we can attempt to
work around those features, or implement them (if they are very desirable) and
contribute back to the open source community.
	\item For any other problem, we will use all the support that the
department provides. We will make sure that the issue is discussed within the
group and we will try to resolve it ourselves. if we find we cannot, then we
will seek additional help.
	\item Each member will make sure that what they are doing is well
documented, so that other members can easily pick up the work. Each member of
the team will also ensure their work is regularly checked in to the git
repository.
\end{itemize}


\section{Legal Issues}
In some cases there is the potential for copyright infringement when dealing
with certain algorithms, this however itself is unlikely, as we will be focusing
on published public algorithms. We will also not be looking to turn a profit
from the use of these algorithms, but to use them for demonstrations of our
developed tools.

There is also an issue of using pieces of open source software, we will make
sure that the use of any such software is documented, with the appropriate
acknowledgements given.

\section{Conclusion}

For our project we aim to create tools that aid in the debugging of Wireless
Sensor Network applications. We plan to implement various predicate detection
algorithms and have the network report to a base station in the case that these
predicates are violated. We also plan to create a new tool that can control
submitting new predicates and visualise the network state, including any
predicate failures.

\clearpage

\appendixpage
\addappheadtotoc
\appendix


\section{References}
\renewcommand{\refname}{\vspace{-1cm}}
\bibliographystyle{myplainnat}
\bibliography{../References/references}

\end{document}
